---
title: "Carmel Valley Wildfire Smoke 'Diurnicity'"
author: "Jonathan Callahan"
date: "2022-10-14"
output: html_document
---

```{r setup, include = FALSE, fig.height = 5, fig.width = 8}
knitr::opts_chunk$set(echo = TRUE)
```

# Background

During the Soberanes Fire in California in 2016 [ref], the air quality monitoring 
station in Carmel Valley showed very high levels of particulate matter (PM2.5)
from wildfire smoke which presented a serious public health risk to local 
residents, workers and firefighters. [ref] During the height of the smoke impacts
in Carmel Valley, relatively calm weather lead to strong diurnal variations in 
air quality as land and sea breezes pushed smoke up and down the valley without
the winds ever being strong enough to clear out the smoke. [ref]

The strong diurnal variation in air quality meant that, even during days with
the worst daily average air quality, there were some hours when outside air 
was in the GOOD or MODERATE categories according to the Air Quality Index (AQI). [ref] 
This time-of-day variation in air quality can be supremely important information
for those living in smoke-impacted areas as it allows them to plan when to hunker
down inside and when to shop, check in on friends, or walk the dog.

This document presents a method for identifying days in an air quality time
series that display strong diurnal cycles during periods of elevated smoke. 
We will create a _diurnicity_ metric that can be applied elsewhere.

Having identified days of interest using the _diurnicity_ metric, we then 
demonstrate how to find the hour of the day at which a maximum value of PM 2.5 
occurred so that we can advise people when to stay indoors and when it is safe
to breathe outdoor air.

# Method

We will use Carmel Valley monitoring data from 2016 to develop an algorithm for
detecting _diurnicity_.

We begin by downloading annual air quality data for 2016 and displaying an 
interactive map of California monitors during the first week of August, 2016.
This map displays monitors colored by the maximum hourly value experienced during
that time period.

The Carmel Valley monitor is located below San Francisco and is colored as
HAZARDOUS.

```{r load_data, echo = FALSE}
library(AirMonitor)
library(MazamaRollUtils)

# ----- Download data ----------------------------------------------------------

airnow_2016 <- airnow_loadAnnual(2016)

# Use interactive map to identify Carmel Valley deviceDeploymentID
airnow_2016 %>%
  monitor_filter(stateCode == "CA") %>%
  monitor_filterDate(20160801, 20160808) %>%
  monitor_dropEmpty() %>%
  monitor_leaflet()

Carmel_Valley <-
  airnow_2016 %>%
  monitor_select("a9572a904a4ed46d_840060530002")
```

## Soberanes Fire

The Soberanes fire began on July 22, 2016 and was not fully extinguished until
October 12, 2016. The bulk of the smoke impacts in the Carmel Valley occurred
in the first three weeks of the fire. [ref]

The following plot shows the diurnal cycle of PM 2.5 measurement with individual
measurements plotted by time-of-day along with averages of all measurements for
each time-of-day. A strong diurnal signal is present with UNHEALTHY conditions
(red on the AQI scale) many mornings and GOOD conditions most evenings.

```{r daily_by_hour_plot, echo = FALSE, warning = FALSE}

# ----- What does a diurnal signal look like -----------------------------------

Carmel_Valley %>%
  monitor_filterDate(20160722, 20160815) %>%
  AirMonitorPlots::monitor_ggDailyByHour_archival(
    title = "Carmel Valley -- Soberanes Fire"
  )
```

## _"Diurnicity"_

A simple metric to capture this diurnal signal during periods of strong
health impacts can be calculated by first converting PM 2.5 values to their
equivalent AQI Level (GOOD = 1, MODERATE = 2, _etc._) and then using rolling 
versions of _max_, _mean_ and _min_ functions to calculate 24-hr averages.

In R pseudocode, this would look like:

```
min <- roll_min(hourly_AQI_level)
mean <- roll_mean(hourly_AQI_level)
max <- roll_max(hourly_AQI_level)

diurnicity <- 10 * mean * (max - min)
```

Each of `min`, `mean` and `max` will vary between AQI levels 1 and 6. The factor 
of 10 is included so that _diurnicity_ will be visible when plotted on top of 
monitoring data.

During the worst days of the Soberanes fire, daily averages were at the 
UNHEALTHY level and the _diurnicity_ would evalute to: 

```
10 * 4 * (6 - 2) = 160
```

This new _diurnicity_ metric can be used to identify time periods of interest
as seen in the following plots. Day-night shading is used to more easily 
identify the diurnal cycle.

A dashed line at `diurnicity = 20` suggests a threshold above which days can
be considered to have a both PM 2.5 levels of concern and a strong diurnal cycle.

```{r diurnicity_plot, echo = FALSE}

# ----- "Diurnicity:  Carmel Valley --------------------------------------------

timezone <- Carmel_Valley$meta$timezone
t <- Carmel_Valley$data$datetime
x <- Carmel_Valley$data[,2]

a <- .bincode(x, breaks = US_AQI$breaks_PM2.5, right = FALSE)

a_max <- roll_max(a, 24, align = "left", na.rm = TRUE)
a_mean <- roll_mean(a, 24, align = "left", na.rm = TRUE)
a_min <- roll_min(a, 24, align = "left", na.rm = TRUE)

diurnicity <- 10 * a_mean * (a_max - a_min)

layout(matrix(seq(1:4), ncol = 2, byrow = TRUE))
par(mar = c(2, 4, 4, 2) + 0.1)

for ( i in seq(0,6,2) ) {

  startdate <- MazamaCoreUtils::parseDatetime(20160701, timezone = timezone) + lubridate::dweeks(i)
  enddate <- startdate + lubridate::dweeks(2)

  Carmel_Valley %>%
    monitor_filterDate(startdate, enddate) %>%
    monitor_timeseriesPlot(shadedNight = TRUE, pch = 15, cex = 0.5, ylim = c(0, 200), xlab = NULL)

  addAQIStackedBar(palette = "deuteranopia")

  # diurnicity ranges goes up to 20+ so multiply by 10 to plot
  lines(t, diurnicity, lwd = 3, col = adjustcolor("coral", 0.5))
  abline(h = 20, lty = "dashed")

  legend(
    "topright",
    legend = "'diurnicity'",
    lwd = 3,
    col = adjustcolor("coral", 0.5)
  )

}

par(mar = c(1, 4, 4, 2) + 0.1)
layout(1)

```

## Hour of PM 2.5 Max during strong _diurnicity_

Our overall goal is to identify those days with strong _diurnicity_ and to 
determine when the maximum PM 2.5 value occurred on those days.

We can identify "diurnal days" using the threshold suggested above. We can then
calculate a rolling 24-hour maximum to identify the local time-of-day.
Combining them, we can identify and then tabulate the time-of-day for all
days with a strong diurnal signal.

In R pseudocode:

```
diurnal_day_mask <- diurnicity > 20
24_hour_mask <- time == roll_max(time)
```

The following plots show how the masks identify the times we are interested in:

```{r time_of_max, echo = FALSE}
# ----- Carmel Valley July-August ----------------------------------------------

Carmel_Valley_JA <-
  Carmel_Valley %>%
  monitor_filterDate(20160701, 20160901, timezone = "America/Los_Angeles")

t_JA <- Carmel_Valley_JA$data$datetime
x_JA <- Carmel_Valley_JA$data[,2]

a_JA <- .bincode(x_JA, breaks = US_AQI$breaks_PM2.5, right = FALSE)

a_JA_max <- roll_max(a_JA, 24, align = "left", na.rm = TRUE)
a_JA_mean <- roll_mean(a_JA, 24, align = "left", na.rm = TRUE)
a_JA_min <- roll_min(a_JA, 24, align = "left", na.rm = TRUE)

diurnicity_JA <- 10 * a_JA_mean * (a_JA_max - a_JA_min)

layout(matrix(seq(3)))
par(mar = c(1, 4, 4, 2) + 0.1)

# Annual w/ cutoff
Carmel_Valley_JA %>%
  monitor_timeseriesPlot(shadedNight = FALSE, pch = 15, cex = 1.0, ylim = c(0, 400), xlab = NULL)

addAQIStackedBar(palette = "deuteranopia")

# diurnicity ranges goes up to 20+ so multiply by 10 to plot
lines(t_JA, diurnicity_JA, lwd = 3, col = adjustcolor("coral", 0.5))
abline(h = 20, lty = "dashed")

# Annual w/ all hours
Carmel_Valley_JA %>%
  monitor_timeseriesPlot(shadedNight = FALSE, pch = 15, cex = 0.8, ylim = c(0,400), xlab = NULL, main = "Times of High Diurnicity")

diurnal_mask <- diurnicity_JA > 20
points(t_JA[diurnal_mask], x_JA[diurnal_mask], pch = 15, cex = 0.8, col = 'red')

addAQIStackedBar(palette = "deuteranopia")

# Annual w/ all maxes
Carmel_Valley_JA %>%
  monitor_timeseriesPlot(shadedNight = FALSE, cex = 0.8, ylim = c(0, 400), xlab = NULL, main = "Hours of Max PM 2.5 during High Diurnicity")

###points(t_JA[diurnal_mask], x_JA[diurnal_mask], pch = 15, cex = 0.8, col = adjustcolor('red', 0.5))

rolling_max_mask <- x_JA == roll_max(x_JA, 24, align = "center", na.rm = TRUE)
diurnal_max_mask <- diurnal_mask & rolling_max_mask

points(t_JA[diurnal_max_mask], x_JA[diurnal_max_mask], pch = 1, cex = 1.0, lwd = 2, col = 'red')

addAQIStackedBar(palette = "deuteranopia")

par(mar = c(1, 4, 4, 2) + 0.1)
layout(1)
```

## Time-of-day

The times identified above can now be tabulated to see when the strong diurnal
cycle typically peaks. In fact, we already know from the very first plot that 
the peak occurs in the morning. But the algorithm we have just developed is very fast
(it uses C++ rollling functions) and could be applied to all monitors over 
multiple years to gain a statistical understanding of different categories of
_diurnicity_.

For Carmel Valley during the Soberanes fire, the hours of maximum PM2.5 during
_diurnicity_ are seen below:

```{r diurnicity_histogram, echo = FALSE}
# ----- What hours are involved ------------------------------------------------

diurnal_max_hour_of_day <-
  t_JA[diurnal_max_mask] %>%
  na.omit() %>%
  lubridate::with_tz(timezone) %>%
  lubridate::hour()

hist(
  diurnal_max_hour_of_day,
  right = FALSE,
  xlim = c(0, 24),
  main = "Hour of Daily Max at Carmel Valley -- Soberanes Fire",
  xlab = "Local time",
  axes = FALSE
)
segments(0, 0, 24, 0)
axis(1, at = c(0, 6, 12, 18, 24), labels = c("midnight", "6 AM", "noon", "6 PM", "midnight"))
axis(2, las = 1)
```

# Next Steps

## Identify all periods with _diurnicity_

With this new metric in hand, it would be straightforward to run through 10 years
worth of data for all AirNow monitoring stations and identify periods of strong
_diurnicity_.

## Determine phase of _diurnicity_

As _diurnicity_ implies an almost sinusoidal signal, we should be able to use
the time of the daily maximum to identify a "phase" fore each trace -- _i.e._ the rough
time of day of the peak. I am imagining we could divide the day up into 3-hour
chunks for this.

## Cluster time series based on strength and phase of _diurnicity_

If we find enough periods with strong _diurnicity_, we could use clustering or
other machine learning techniques to potentially categorize different types of
locations, weather patterns or fire behavior.


