---
title: "Carmel Valley Wildfire Smoke 'Diurnicity'"
author: "Jonathan Callahan"
date: "2021-12-14"
output: html_document
---

```{r setup, include = FALSE, fig.height = 5, fig.width = 8}
knitr::opts_chunk$set(echo = TRUE)
```

# Background

Recent conversations brought up concerns about using midnight as the
day-boundary when calculating daily average PM 2.5 levels. In locations where
the daily max occurs at night, daily average PM2.5 levels might be reduced
by splitting the highest values between two days. (This is likely to occur
in valley locations that act as pour points for down-valley drainage of
smoke during nighttime smoldering.)

This document presents a method for identifying days in an air quality time
series that display strong diurnal cycles during periods of elevated smoke. 
We will call this _diurnicity_ for the time being.

Having identified days of interest, we then find the hour of the day at which
a maximum value of PM 2.5 occurred so that we can assess whether a midnight
day boundary will split high PM 2.5 values across two days. In a perfect world,
our day boundaries would be shifted 12 hours from the time of each daily max.

# Method

During the Soberanes Fire in 2016, the air quality monitoring station in the 
Carmel Valley showed very high levels of smoke and a very strong diurnal
signal affected largely by land and sea breezes during otherwise calm weather.

We will use the Carmel Valley monitoring data to develop an algorithm for
detecting _diurnicity_.

```{r load_data, echo = FALSE}
library(AirMonitor)
library(MazamaRollUtils)

# ----- Download data ----------------------------------------------------------

AirNow_2016 <-
  PWFSLSmoke::airnow_loadAnnual(2016) %>%
  monitor_fromPWFSLSmoke()
```

## Sobranes Fire

The Soberanes fire began on July 22, 2016 and was not fully extinguished until
October 12, 2016. The bulk of the smoke impacts in the Carmel Valley occurred
in the first three weeks of the fire.

The following plot shows the diurnal cycle of PM 2.5 measurement with individual
measurements plotted by time-of-day along with averages of all measurements for
each time-of-day. A strong diurnal signal is present with UNHEALTHY conditions
(red on the AQI scale) many mornings and GOOD conditions most evenings.

```{r daily_by_hour_plot, echo = FALSE, warning = FALSE}

# ----- What does a diurnal signal look like -----------------------------------

AirMonitor::Carmel_Valley %>%
  monitor_toPWFSLSmoke() %>%
  AirMonitorPlots::monitor_ggDailyByHour_archival(
    title = "Carmel Valley -- Sobranes Fire"
  )
```

## _"Diurnicity"_

A simple metric to capture this diurnal signal during periods of strong
health impacts can be calculated by first converting  PM 2.5 values to their
equivalent AQI Level (GOOD = 1, MODERATE = 2, _etc._) and then using rolling 
versions of _max_, _mean_ and _min_ functions.

In vectorized pseudocode this would look like:

```
min <- roll_min(hourly_AQI_level)
mean <- roll_mean(hourly_AQI_level)
max <- roll_max(hourly_AQI_level)

diurnicity <- 10 * mean * (max - min)
```

Each of `min`, `mean` and `max` will vary between 1 and 6. The factor of 10 is
included so that _diurnicity_ will be visible when plotted on top of monitoring
data.

During the worst days of the Soberanes fire, daily averages were at the 
UNHEALTHY level and the _diurnicity_ would evalute to: 

```
10 * 4 * (6 - 2) = 160
```

This new _diurnicity_ metric can be used to identify time periods of interest
as seen in the following plots. Day-night shading is used to more easily 
identify the diurnal cycle.

A dashed line at `diurnicity = 20` suggests a threshold above which days can
be considered to have a both PM 2.5 levels of concern and a strong diurnal cycle.

```{r diurnicity_plot, echo = FALSE}

# ----- "Diurnicity:  Carmel Valley --------------------------------------------

Carmel_Valley <-
  AirNow_2016 %>%
  monitor_filter(AQSID == "060530002")

timezone <- Carmel_Valley$meta$timezone
t <- Carmel_Valley$data$datetime
x <- Carmel_Valley$data[,2]

a <- .bincode(x, breaks = US_AQI$breaks_PM2.5, right = FALSE)

a_max <- roll_max(a, 24, align = "left", na.rm = TRUE)
a_mean <- roll_mean(a, 24, align = "left", na.rm = TRUE)
a_min <- roll_min(a, 24, align = "left", na.rm = TRUE)

diurnicity <- 10 * a_mean * (a_max - a_min)

layout(matrix(seq(1:4), ncol = 2, byrow = TRUE))
par(mar = c(2, 4, 4, 2) + 0.1)

for ( i in seq(0,6,2) ) {

  startdate <- MazamaCoreUtils::parseDatetime(20160701, timezone = timezone) + lubridate::dweeks(i)
  enddate <- startdate + lubridate::dweeks(2)

  Carmel_Valley %>%
    monitor_filterDate(startdate, enddate) %>%
    monitor_timeseriesPlot(shadedNight = TRUE, pch = 15, cex = 0.5, ylim = c(0, 200), xlab = NULL)

  addAQIStackedBar(palette = "deuteranopia")

  # diurnicity ranges goes up to 20+ so multiply by 10 to plot
  lines(t, diurnicity, lwd = 3, col = adjustcolor("coral", 0.5))
  abline(h = 20, lty = "dashed")

  legend(
    "topright",
    legend = "'diurnicity'",
    lwd = 3,
    col = adjustcolor("coral", 0.5)
  )

}

par(mar = c(1, 4, 4, 2) + 0.1)
layout(1)

```

## Hour of PM 2.5 Max during strong _diurnicity_

Our overall goal is to identify those days with strong _diurnicity_ and to 
determine when the maximum PM 2.5 value occurred on those days.

We can identify "diurnal days" using the threshold suggested above. We can then
calculate a rolling 24-hour maximum to identify the local time-of-day.
Combining them, we can identify and then tabulate the time-of-day for all
days with a strong diurnal signal.

In vectorized pseudocode:

```
diurnal_day_mask <- diurnicity > 20
24_hour_mask <- time == roll_max(time)
```

The following plots show how the masks identify the times we are interested in:

```{r time_of_max, echo = FALSE}
# ----- Carmel Valley July-August ----------------------------------------------

Carmel_Valley_JA <-
  Carmel_Valley %>%
  monitor_filterDate(20160701, 20160901, timezone = "America/Los_Angeles")

t_JA <- Carmel_Valley_JA$data$datetime
x_JA <- Carmel_Valley_JA$data[,2]

a_JA <- .bincode(x_JA, breaks = US_AQI$breaks_PM2.5, right = FALSE)

a_JA_max <- roll_max(a_JA, 24, align = "left", na.rm = TRUE)
a_JA_mean <- roll_mean(a_JA, 24, align = "left", na.rm = TRUE)
a_JA_min <- roll_min(a_JA, 24, align = "left", na.rm = TRUE)

diurnicity_JA <- 10 * a_JA_mean * (a_JA_max - a_JA_min)

layout(matrix(seq(3)))
par(mar = c(1, 4, 4, 2) + 0.1)

# Annual w/ cutoff
Carmel_Valley_JA %>%
  monitor_timeseriesPlot(shadedNight = FALSE, pch = 15, cex = 1.0, ylim = c(0, 400), xlab = NULL)

addAQIStackedBar(palette = "deuteranopia")

# diurnicity ranges goes up to 20+ so multiply by 10 to plot
lines(t_JA, diurnicity_JA, lwd = 3, col = adjustcolor("coral", 0.5))
abline(h = 20, lty = "dashed")

# Annual w/ all hours
Carmel_Valley_JA %>%
  monitor_timeseriesPlot(shadedNight = FALSE, pch = 15, cex = 0.8, ylim = c(0,400), xlab = NULL, main = "Times of High Diurnicity")

diurnal_mask <- diurnicity_JA > 20
points(t_JA[diurnal_mask], x_JA[diurnal_mask], pch = 15, cex = 0.8, col = 'red')

addAQIStackedBar(palette = "deuteranopia")

# Annual w/ all maxes
Carmel_Valley_JA %>%
  monitor_timeseriesPlot(shadedNight = FALSE, cex = 0.8, ylim = c(0, 400), xlab = NULL, main = "Hours of Max PM 2.5 during High Diurnicity")

###points(t_JA[diurnal_mask], x_JA[diurnal_mask], pch = 15, cex = 0.8, col = adjustcolor('red', 0.5))

rolling_max_mask <- x_JA == roll_max(x_JA, 24, align = "center", na.rm = TRUE)
diurnal_max_mask <- diurnal_mask & rolling_max_mask

points(t_JA[diurnal_max_mask], x_JA[diurnal_max_mask], pch = 1, cex = 1.0, lwd = 2, col = 'red')

addAQIStackedBar(palette = "deuteranopia")

par(mar = c(1, 4, 4, 2) + 0.1)
layout(1)
```

## Time-of-day

The times identified above can now be tabulated to see when the strong diurnal
cycle typically peaks. In fact, we already know from the very first plot that 
the peak occurs in the morning. But the algorithm we have just developed is very fast
(it uses C++ rollling functions) and could be applied to all monitors over 
multiple years to gain a statistical understanding of different categories of
_diurnicity_.

For Carmel Valley during the Soberanes fire, the hours of maximum PM2.5 during
_diurnicity_ are seen below:

```{r diurnicity_histogram, echo = FALSE}
# ----- What hours are involved ------------------------------------------------

diurnal_max_hour_of_day <-
  t_JA[diurnal_max_mask] %>%
  na.omit() %>%
  lubridate::with_tz(timezone) %>%
  lubridate::hour()

hist(
  diurnal_max_hour_of_day,
  right = FALSE,
  xlim = c(0, 24),
  main = "Hour of Daily Max at Carmel Valley -- Soberanes Fire",
  xlab = "Local time",
  axes = FALSE
)
segments(0, 0, 24, 0)
axis(1, at = c(0, 6, 12, 18, 24), labels = c("midnight", "6 AM", "noon", "6 PM", "midnight"))
axis(2, las = 1)
```

# Next Steps

## Identify all periods with _diurnicity_

With this new metric in hand, it would be straightforward to run through 10 years
worth of data for all AirNow monitoring stations and identify periods of strong
_diurnicity_.

## Determine phase of _diurnicity_

As _diurnicity_ implies an almost sinusoidal signal, we should be able to use
the time of the daily maximum to identify a "phase" fore each trace -- _i.e._ the rough
time of day of the peak. I am imagining we could divide the day up into 3-hour
chunks for this.

## Cluster time series based on strength and phase of _diurnicity_

If we find enough periods with strong _diurnicity_, we could use clustering or
other machine learning techniques to potentially categorize different types of
locations or different types of fire behavior.

## Answer the question of best day-boundary

A thorough analysis of all monitoring time series in the past decade should 
give us a very good handle on whether midnight or some other time would make
a good  day boundnary.

