---
title: "Introduction to AirMonitor"
author: "Mazama Science"
date: "2020-02-12"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to AirMonitor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 7, fig.height = 5)
```

## Installation

Install from CRAN with:

`install.packages('AirMonitor')`

Install the latest version from GitHub with:

`devtools::install_github('mazamascience/AirMonitor')`

## Available data

The USFS AirFire group regularly processes monitoring data in support of their
various operational tools. Pre-processed, harmonized and QC'ed data files can
be loaded with the following functions:

* `~_load()` -- load data based on a start- and end-time
* `~loadAnnual()` -- load a year's worth of data
* `~loadDaily()` -- load the most recent 45 days of data (updated once per day)
* `~loadLatest()` -- load the most recent 10 days of data (updated every hour)

Data archives go back over 10 years depending on the data source.

## Data pipelines

We encourage people to embrace "data pipeline" style coding as encouraged by
**dplyr** and related packages. The special `%>%` operator uses the output
of one function as the first argument of the next function, thus allowing for 
easy "chaining" of results.

With only a few exceptions, all the `monitor_` functions accept a _mts_monitor_ 
object as their first argument and generate a _mts_monitor_ object as a result 
so they can be chained together.

## A first example

```{r library, echo = FALSE}
suppressPackageStartupMessages({
  library(AirMonitor)
  
  Camp_Fire <- Camp_Fire
})
```

Let's say we are interested in the impact of smoke from the 2018 
[Camp Fire](https://en.wikipedia.org/wiki/Camp_Fire_(2018)) in the Sacramento
area.

We would begin by creating a `Camp_Fire` _mts_monitor_ object that has all the
monitors in California for the period of interest:

```
Camp_Fire <-
  monitor_loadAnnual(2018) %>%
  monitor_filter(stateCode == 'CA') %>%
  monitor_filterDate(
    startdate = 20181108,
    enddate = 20181123,
    timezone = "America/Los_Angeles"
  ) %>%
  monitor_dropEmpty()
```

We can display these monitors (colored by maximum PM2.5 value) in an interactive
map, zoom in and click on the monitor in downtown Sacramento to get it's 
`deviceDeploymentID`:

```{r Sacramento_2}
monitor_leaflet(Camp_Fire)
```

We can use this `deviceDeploymentID` to create a _mts_monitor_ object for this 
single monitor and take a look at a timeseries plot:

```{r Sacramento_3}
Sacramento <-
  Camp_Fire %>%
  monitor_select("8ca91d2521b701d4_060670010")

Sacramento %>%
  monitor_timeseriesPlot(
    shadedNight = TRUE,
    addAQI = TRUE
  )

addAQILegend(cex = 0.8)
```

Next, we can use this specific location to create a _mts_monitor_ object
containing all monitors within 50 kilometers (31 miles) of Sacramento:

```{r Sacramento_4}
Sacramento_area <-
  Camp_Fire %>%
  monitor_filterByDistance(
    longitude = Sacramento$meta$longitude,
    latitude = Sacramento$meta$latitude,
    radius = 50000
  )

monitor_leaflet(Sacramento_area)
```

We can display the hourly data for all the monitors and add day/night shading:

```{r Sacramento_5}
Sacramento_area %>%
  monitor_timeseriesPlot(
    shadedNight = TRUE,
    addAQI = TRUE,
    main = "Wildfire Smoke within 30 miles of Sacramento"
  )
addAQILegend(lwd = 1, pch = NA, bg = "white")
```

We can average together all the monitors and create a local-time daily average
for the Sacramento area:

```{r Sacramento_6}
Sacramento_area %>%
  monitor_collapse(
    deviceID = "Sacramento_area"
  ) %>%
  monitor_dailyStatistic() %>%
  monitor_getData()
```

Alternatively, we can plot the daily averages:

```{r Sacramento_7}
Sacramento_area %>%
  monitor_collapse() %>%
  monitor_dailyBarplot(
    main = "Daily Average PM2.5 in the Sacramento Area"
  )
addAQILegend(pch = 15, bg = "white")
```

----

Best of luck analyzing your local air quality data!

