---
title: "AIRSIS Data Extremes 2021"
author: "Jonathan Callahan"
date: "2022-09-08"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(AirMonitor)

cutoff <- 750
roc_cutoff <- 250

airsis <- 
  airsis_loadAnnual(2021) %>%
  monitor_filterDate(20210701, 20211001, timezone = "UTC") %>%
  monitor_replaceValues(data > 4000, NA)
```

This mini-report takes a look at the data values from temporary
monitors whose data is handled by the AIRSIS (aka Oceaneering).

We compare data values from these monitors against suggested QC chcecks for:

* Max severe (invalid) range = `r cutoff` ug/m3
* Rate of change = `r roc_cutoff` ug/m3/hr

# AIRSIS Data

During the fall of 2021, 132 monitors were deployed whose data was transmitted via
satellite to AIRSIS and subsequently processed by AirFire data ingest
scripts. The following plots show the values of PM2.5 for a three month period when wildfire
smoke was impacting communities.

```{r all_data_timeseries_plot, echo = FALSE}
par(mar = c(3, 4, 6, 2) + .1)
monitor_timeseriesPlot(airsis, xpd = NA, main = "2021 PM2.5 Values")
abline(h = cutoff, col = adjustcolor("gray50", 0.6), lwd = 3)
xlo <- par("usr")[1]
xhi <- par("usr")[2]
text(xlo, cutoff+25, sprintf("%d cutoff", cutoff), pos = 4, adj = c(0, 1))
par(mar = c(5, 4, 4, 2) + .1)

par(mar = c(3, 4, 6, 2) + .1)
monitor_timeseriesPlot(airsis, ylim = c(0, 1000), main = "2021 PM2.5 Values (zoomed in)")
abline(h = cutoff, col = adjustcolor("gray50", 0.6), lwd = 3)
xlo <- par("usr")[1]
xhi <- par("usr")[2]
text(xlo, cutoff+25, sprintf("%d cutoff", cutoff), pos = 4, adj = c(0, 1))
par(mar = c(5, 4, 4, 2) + .1)
```

## High Value Cutoff = `r cutoff`

AirNow uses automated QC to invalidate any values above a "Max Severe" level.
The default level for this high value cutoff is `r cutoff` ug/m3.

Below we see two histograms, one with all of the values below the default 
"high value" cutoff of `r cutoff` ug/m3 and a second one with all the values above the cutoff.

```{r all_data_histograms, echo = FALSE}
layout(matrix(seq(2)))
par(mar = c(3,4,2,2) + .1)
data <- airsis$data[,-1] %>% as.matrix() %>% as.numeric()
hist(data[data < cutoff], breaks = seq(0, cutoff, 5), main = sprintf("Histogram of data < %d ug/m3", cutoff))
data <- airsis$data[,-1] %>% as.matrix() %>% as.numeric()
hist(data[data >= cutoff], breaks = seq(cutoff, 2500, by = 50), main = sprintf("Histogram of data >= %d ug/m3", cutoff))
par(mar = c(5,4,4,2) + .1)
layout(1)
```

**Monitors with the highest values**

```{r high_value_IDs, echo = FALSE}
high_value_IDs <-
  airsis %>%
  monitor_replaceValues(data < cutoff, NA) %>%
  monitor_dropEmpty() %>%
  monitor_getMeta() %>%
  dplyr::pull(deviceDeploymentID)

airsis_high <- monitor_select(airsis, high_value_IDs)
```

`r length(high_value_IDs)` monitors in this time period had hourly PM2.5 values above `r cutoff` ug/m3.

```{r high_value_timeseries, echo = FALSE}
airsis_high %>% monitor_timeseriesPlot(type = 'l', main = "Monitors with high values")

abline(h = cutoff, col = adjustcolor("gray50", 0.6), lwd = 3)

xlo <- par("usr")[1]
xhi <- par("usr")[2]
text(xlo, cutoff+25, sprintf("%d cutoff", cutoff), pos = 4, adj = c(0, 1))
```

```{r high_value_timeseries_1, echo = FALSE}
airsis_high %>%
  monitor_filterDate(20210724, 20210806, timezone = "UTC") %>%
  monitor_timeseriesPlot(type = 'l', main = "Monitors with high values")
  
abline(h = cutoff, col = adjustcolor("gray50", 0.6), lwd = 3)

xlo <- par("usr")[1]
xhi <- par("usr")[2]
text(xlo, cutoff+25, sprintf("%d cutoff", cutoff), pos = 4, adj = c(0, 1))
```

```{r high_value_timeseries_2, echo = FALSE}
airsis_high %>%
  monitor_filterDate(20210810, 20210824, timezone = "UTC") %>%
  monitor_timeseriesPlot(type = 'l', main = "Monitors with high values")
  
abline(h = cutoff, col = adjustcolor("gray50", 0.6), lwd = 3)

xlo <- par("usr")[1]
xhi <- par("usr")[2]
text(xlo, cutoff+25, sprintf("%d cutoff", cutoff), pos = 4, adj = c(0, 1))
```

```{r high_value_timeseries_3, echo = FALSE}
airsis_high %>%
  monitor_filterDate(20210901, 20210915, timezone = "UTC") %>%
  monitor_timeseriesPlot(type = 'l', main = "Monitors with high values")
  
abline(h = cutoff, col = adjustcolor("gray50", 0.6), lwd = 3)

xlo <- par("usr")[1]
xhi <- par("usr")[2]
text(xlo, cutoff+25, sprintf("%d cutoff", cutoff), pos = 4, adj = c(0, 1))
```

## Rate of Change Cutoff = `r roc_cutoff` ug/m3/hr

AirNow QC also applies a rate-of-change cutoff of `r roc_cutoff` ug/m3/hr to flag suspect data.
This cutoff may also be too low for monitors deployed near wildfires.

The histograms of rate-of-change for this time period are below. There are many
rate-of-change values above `r roc_cutoff` ug/m3/hr in the monitoring data.

```{r rate_of_change_histograms, echo = FALSE}
dataBrick <- airsis$data[,-1]
data <-
  dataBrick %>%
  lapply(diff) %>%
  unlist() %>%
  as.numeric() #%>%
  #abs()
  
layout(matrix(seq(2)))
par(mar = c(3,4,2,2) + .1)
hist(data[abs(data) < roc_cutoff], breaks = seq(-roc_cutoff, roc_cutoff, 5), main = sprintf("Histogram of rate-of-change < %d ug/m3/hr", roc_cutoff))
hist(data[abs(data) >= roc_cutoff & abs(data) <= 1000], breaks = seq(-1000,1000,roc_cutoff), main = sprintf("Histogram of rate-of-change >= %d ug/m3/hr", roc_cutoff))
par(mar = c(5,4,4,2) + .1)
layout(1)
```

**Monitors with high rate-of-change values**

We can view those high rate-of-change values in context with the following series
of timeseries plots. For the vast majority of cases, these high rate-of-change
values are associated with smoke in an environment with strong diurnal 
variation in PM2.5 levels. All of these values in red are associated with
high rate-of-change and almost all appear to be valid.

```{r high_value_roc, echo = FALSE}
# Determine number of monitors in high group
a <- dataBrick %>% lapply(diff)
b <- lapply(a, function(x) { any(x >= roc_cutoff, na.rm = TRUE) })
c <- unlist(b)

high_roc_IDs <- names(c[c == TRUE]) # any high roc values

high_roc <- 
  airsis %>% 
  monitor_select(high_roc_IDs) %>%
  monitor_dropEmpty()

my_mask_FUN <- function(x) {
  diff_vector <- c(as.numeric(NA), diff(x))
  mask <- diff_vector < roc_cutoff
  mask[is.na(mask)] <- TRUE
  x[mask] <- NA
  return(x)
}

masked_high_roc <- high_roc %>% monitor_mutate(my_mask_FUN)
```

```{r roc_plots, echo = FALSE}
high_roc %>%
  monitor_filterDate(20210715, 20210915, timezone = "UTC") %>%
  monitor_timeseriesPlot(cex = 0.6, shadedNight = TRUE, main = sprintf("Monitors with Rate-of-Change >= %d ug/m3/hr", roc_cutoff))
monitor_timeseriesPlot(high_roc, type = 'l', lwd = 0.5, add = TRUE)
monitor_timeseriesPlot(masked_high_roc, cex = 0.6, col = "red", add = TRUE, opacity = 1)
```

```{r roc_plots_1, echo = FALSE}
high_roc %>%
  monitor_filterDate(20210724, 20210806, timezone = "UTC") %>%
  monitor_timeseriesPlot(cex = 0.6, shadedNight = TRUE, main = sprintf("Monitors with Rate-of-Change >= %d ug/m3/hr", roc_cutoff))
monitor_timeseriesPlot(high_roc, type = 'l', lwd = 0.5, add = TRUE)
monitor_timeseriesPlot(masked_high_roc, cex = 0.6, col = "red", add = TRUE, opacity = 1)
```

```{r roc_plots_2, echo = FALSE}
high_roc %>%
  monitor_filterDate(20210810, 20210824, timezone = "UTC") %>%
  monitor_timeseriesPlot(cex = 0.6, shadedNight = TRUE, main = sprintf("Monitors with Rate-of-Change >= %d ug/m3/hr", roc_cutoff))
monitor_timeseriesPlot(high_roc, type = 'l', lwd = 0.5, add = TRUE)
monitor_timeseriesPlot(masked_high_roc, cex = 0.6, col = "red", add = TRUE, opacity = 1)
```

```{r roc_plots_3, echo = FALSE}
high_roc %>%
  monitor_filterDate(20210901, 20210915, timezone = "UTC") %>%
  monitor_timeseriesPlot(cex = 0.6, shadedNight = TRUE, main = sprintf("Monitors with Rate-of-Change >= %d ug/m3/hr", roc_cutoff))
monitor_timeseriesPlot(high_roc, type = 'l', lwd = 0.5, add = TRUE)
monitor_timeseriesPlot(masked_high_roc, cex = 0.6, col = "red", add = TRUE, opacity = 1)
```

----


