---
title: "WRCC Data Extremes 2021"
author: "Jonathan Callahan"
date: "2022-09-07"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(AirMonitor)

wrcc <- 
  wrcc_loadAnnual(2021) %>%
  monitor_filterDate(20210701, 20211001, timezone = "UTC") %>%
  monitor_replaceValues(data > 4000, NA)
```

This mini-report takes a look at the data values from USFS deployed temporary
monitors whose data is handled by the Western
Regional Climate Center (WRCC) at the Desert Research Institute (DRI).

We compare data values from these monitors against the AirNow default QC chcecks
for:

* Max severe (invalid) range: 175 ug/m3
* Rate of change: 40 ug/m3

# WRCC Data

During the fall of 2021, 49 monitors were deployed whose data was transmitted via
GOES satellite to WRCC and subsequently processed by AirFire data ingest
scripts. The following plots show the values of PM2.5 for a three month period when wildfire
smoke was impacting communities.

```{r all_data_timeseries_plot, echo = FALSE}
par(mar = c(3, 4, 6, 2) + .1)
monitor_timeseriesPlot(wrcc, addAQI = TRUE, xpd = NA, main = "2021 PM2.5 Values")
par(mar = c(5, 4, 4, 2) + .1)
addAQILegend("topleft", cex = 0.8)

par(mar = c(3, 4, 6, 2) + .1)
monitor_timeseriesPlot(wrcc, ylim = c(0, 300), addAQI = TRUE, main = "2021 PM2.5 Values (zoomed in)")
par(mar = c(5, 4, 4, 2) + .1)
addAQILegend("topleft", cex = 0.8)
```

## High Value Cutoff = 175

AirNow uses automated QC to invalidate any values above a "Max Severe" level.
The default level for this high value cutoff is 175 ug/m3.

Below we see two histograms, one with all of the values below the default 
"high value" cutoff of 175 ug/m3 and a second one with all the values above the cutoff.

```{r all_data_histograms, echo = FALSE}
layout(matrix(seq(2)))
par(mar = c(3,4,2,2) + .1)
data <- wrcc$data[,-1] %>% as.matrix() %>% as.numeric()
hist(data[data < 175], breaks = seq(0, 175, 5), main = "Histogram of data < 175 ug/m3")
data <- wrcc$data[,-1] %>% as.matrix() %>% as.numeric()
hist(data[data >= 175], breaks = seq(175, 1400, by = 25), main = "Histogram of data >= 175 ug/m3")
par(mar = c(5,4,4,2) + .1)
layout(1)
```

**Monitors with the highest values**

Three monitors in this time period had hourly PM2.5 values above 175 ug/m3. Of
the three, only the monitor named "Smoke USFS R1-52" appears to have an _invalid_
high value while the monitor named "Smoke USFS 3016" has many valid high values.

```{r high_value_timeseries, echo = FALSE}
high_value_IDs <-
  wrcc %>%
  monitor_replaceValues(data < 175, NA) %>%
  monitor_dropEmpty() %>%
  monitor_getMeta() %>%
  dplyr::pull(deviceDeploymentID)

wrcc_high <- monitor_select(wrcc, high_value_IDs)

# fullAQSID is required in the current version of AirMonitorPlots
wrcc_high$meta$fullAQSID <- wrcc_high$meta$deviceDeploymentID
# locationName is currently just the monitor name and is repeated.
wrcc_high$meta$locationName <- wrcc_high$meta$locationID

wrcc %>% 
  monitor_select(high_value_IDs[3]) %>%
  monitor_filterDate(20210715, 20210915, timezone = "UTC") %>%
  monitor_timeseriesPlot(type = 'l', col = 'black', main = "Monitors with high values")

wrcc %>% 
  monitor_select(high_value_IDs[1]) %>%
  monitor_timeseriesPlot(type = 'l', col = 'dodgerblue', add = TRUE)

wrcc %>% 
  monitor_select(high_value_IDs[2]) %>%
  monitor_timeseriesPlot(type = 'l', col = 'dodgerblue', add = TRUE)

wrcc %>% 
  monitor_select(high_value_IDs[4]) %>%
  monitor_timeseriesPlot(type = 'l', col = 'salmon', add = TRUE)

abline(h = 175, col = adjustcolor("gray50", 0.6), lwd = 3)

xlo <- par("usr")[1]
xhi <- par("usr")[2]
text(xlo, 200, "175 cutoff", pos = 4, adj = c(0, 1))

legend(
  "topleft",
  legend = c("Smoke #86", "Smoke USFS R1-52", "Smoke USFS 3016"),
  lwd = c(1,1,1),
  col = c("dodgerblue", "salmon", "black")
)
```

```{r high_value_timeseries_1, echo = FALSE}
wrcc_high %>%
  monitor_filterDate(20210724, 20210806, timezone = "UTC") %>%
  monitor_timeseriesPlot(type = 'l', col = 'transparent', main = "Monitors with high values")
  
wrcc %>% 
  monitor_select(high_value_IDs[3]) %>%
  monitor_timeseriesPlot(type = 'l', col = 'black', add = TRUE)

wrcc %>% 
  monitor_select(high_value_IDs[1]) %>%
  monitor_timeseriesPlot(type = 'l', col = 'dodgerblue', add = TRUE)

wrcc %>% 
  monitor_select(high_value_IDs[2]) %>%
  monitor_timeseriesPlot(type = 'l', col = 'dodgerblue', add = TRUE)

wrcc %>% 
  monitor_select(high_value_IDs[4]) %>%
  monitor_timeseriesPlot(type = 'l', col = 'salmon', add = TRUE)

abline(h = 175, col = adjustcolor("gray50", 0.6), lwd = 3)

xlo <- par("usr")[1]
xhi <- par("usr")[2]
text(xlo, 200, "175 cutoff", pos = 4, adj = c(0, 1))

legend(
  "topleft",
  legend = c("Smoke #86", "Smoke USFS R1-52", "Smoke USFS 3016"),
  lwd = c(1,1,1),
  col = c("dodgerblue", "salmon", "black")
)
```

```{r high_value_timeseries_2, echo = FALSE}
wrcc_high %>%
  monitor_filterDate(20210810, 20210824, timezone = "UTC") %>%
  monitor_timeseriesPlot(type = 'l', col = 'transparent', main = "Monitors with high values")
  
wrcc %>% 
  monitor_select(high_value_IDs[3]) %>%
  monitor_timeseriesPlot(type = 'l', col = 'black', add = TRUE)

wrcc %>% 
  monitor_select(high_value_IDs[1]) %>%
  monitor_timeseriesPlot(type = 'l', col = 'dodgerblue', add = TRUE)

wrcc %>% 
  monitor_select(high_value_IDs[2]) %>%
  monitor_timeseriesPlot(type = 'l', col = 'dodgerblue', add = TRUE)

wrcc %>% 
  monitor_select(high_value_IDs[4]) %>%
  monitor_timeseriesPlot(type = 'l', col = 'salmon', add = TRUE)

abline(h = 175, col = adjustcolor("gray50", 0.6), lwd = 3)

xlo <- par("usr")[1]
xhi <- par("usr")[2]
text(xlo, 200, "175 cutoff", pos = 4, adj = c(0, 1))

legend(
  "topright",
  legend = c("Smoke #86", "Smoke USFS R1-52", "Smoke USFS 3016"),
  lwd = c(1,1,1),
  col = c("dodgerblue", "salmon", "black")
)
```

```{r high_value_timeseries_3, echo = FALSE}
wrcc_high %>%
  monitor_filterDate(20210901, 20210915, timezone = "UTC") %>%
  monitor_timeseriesPlot(type = 'l', col = 'transparent', main = "Monitors with high values")
  
wrcc %>% 
  monitor_select(high_value_IDs[3]) %>%
  monitor_timeseriesPlot(type = 'l', col = 'black', add = TRUE)

wrcc %>% 
  monitor_select(high_value_IDs[1]) %>%
  monitor_timeseriesPlot(type = 'l', col = 'dodgerblue', add = TRUE)

wrcc %>% 
  monitor_select(high_value_IDs[2]) %>%
  monitor_timeseriesPlot(type = 'l', col = 'dodgerblue', add = TRUE)

wrcc %>% 
  monitor_select(high_value_IDs[4]) %>%
  monitor_timeseriesPlot(type = 'l', col = 'salmon', add = TRUE)

abline(h = 175, col = adjustcolor("gray50", 0.6), lwd = 3)

xlo <- par("usr")[1]
xhi <- par("usr")[2]
text(xlo, 200, "175 cutoff", pos = 4, adj = c(0, 1))

legend(
  "topleft",
  legend = c("Smoke #86", "Smoke USFS R1-52", "Smoke USFS 3016"),
  lwd = c(1,1,1),
  col = c("dodgerblue", "salmon", "black")
)
```

## Rate of Change Cutoff = 40 ug/m3/hr

AirNow QC also applies a rate-of-change cutoff of 40 ug/m3/hr to flag suspect data.
This cutoff may also be too low for monitors deployed near wildfires.

The histograms of rate-of-change for this time period are below. There is a 
reasonable sprinkling of rate-of-change values above 40 ug/m3/hr associated with
7 separate monitors.

```{r rate_of_change_histograms, echo = FALSE}
dataBrick <- wrcc$data[,-1]
data <-
  dataBrick %>%
  lapply(diff) %>%
  unlist() %>%
  as.numeric() #%>%
  #abs()
  
layout(matrix(seq(2)))
par(mar = c(3,4,2,2) + .1)
hist(data[abs(data) < 40], breaks = seq(-40, 40, 5), main = "Histogram of rate-of-change < 40 ug/m3/hr")
hist(data[abs(data) >= 40 & abs(data) <= 1000], breaks = seq(-1000,1000,40), main = "Histogram of rate-of-change >= 40 ug/m3/hr")
par(mar = c(5,4,4,2) + .1)
layout(1)
```

**Monitors with high rate-of-change values**

We can view those high rate-of-change values in context with the following series
of timeseries plots. For the vast majority of cases, these high rate-of-change
values are associated with smoke in an environment with strong diurnal 
variation in PM2.5 levels. All of these values in red are associated with
high rate-of-change and almost all appear to be valid.

```{r high_value_roc, echo = FALSE}
# Determine number of monitors in high group
a <- dataBrick %>% lapply(diff)
b <- lapply(a, function(x) { any(x >= 40, na.rm = TRUE) })
c <- unlist(b)

high_roc_IDs <- names(c[c == TRUE]) # any high roc values

high_roc <- 
  wrcc %>% 
  monitor_select(high_roc_IDs) %>%
  monitor_dropEmpty()

my_mask_FUN <- function(x) {
  diff_vector <- c(as.numeric(NA), diff(x))
  mask <- diff_vector < 40
  mask[is.na(mask)] <- TRUE
  x[mask] <- NA
  return(x)
}

masked_high_roc <- high_roc %>% monitor_mutate(my_mask_FUN)
```

```{r roc_plots, echo = FALSE}
high_roc %>%
  monitor_filterDate(20210715, 20210915, timezone = "UTC") %>%
  monitor_timeseriesPlot(cex = 0.6, shadedNight = TRUE, main = "Monitors with Rate-of-Change >= 40 ug/m3/hr")
monitor_timeseriesPlot(high_roc, type = 'l', lwd = 0.5, add = TRUE)
monitor_timeseriesPlot(masked_high_roc, cex = 0.6, col = "red", add = TRUE)
```

```{r roc_plots_1, echo = FALSE}
high_roc %>%
  monitor_filterDate(20210724, 20210806, timezone = "UTC") %>%
  monitor_timeseriesPlot(cex = 0.6, shadedNight = TRUE, main = "Monitors with Rate-of-Change >= 40 ug/m3/hr")
monitor_timeseriesPlot(high_roc, type = 'l', lwd = 0.5, add = TRUE)
monitor_timeseriesPlot(masked_high_roc, cex = 0.6, col = "red", add = TRUE)
```

```{r roc_plots_2, echo = FALSE}
high_roc %>%
  monitor_filterDate(20210810, 20210824, timezone = "UTC") %>%
  monitor_timeseriesPlot(cex = 0.6, shadedNight = TRUE, main = "Monitors with Rate-of-Change >= 40 ug/m3/hr")
monitor_timeseriesPlot(high_roc, type = 'l', lwd = 0.5, add = TRUE)
monitor_timeseriesPlot(masked_high_roc, cex = 0.6, col = "red", add = TRUE)
```

```{r roc_plots_3, echo = FALSE}
high_roc %>%
  monitor_filterDate(20210901, 20210915, timezone = "UTC") %>%
  monitor_timeseriesPlot(cex = 0.6, shadedNight = TRUE, main = "Monitors with Rate-of-Change >= 40 ug/m3/hr")
monitor_timeseriesPlot(high_roc, type = 'l', lwd = 0.5, add = TRUE)
monitor_timeseriesPlot(masked_high_roc, cex = 0.6, col = "red", add = TRUE)
```

----


