---
title: "NowCast"
author: "Rex Thompson and Jonathan Callahan"
date: "2022-10-31"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NowCast}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 5, comment = NA)
options(width = 105)
```

This vignette documents the `monitor_nowcast()` function, which converts a 
_mts_monitor_ object's values to NowCast values. We provide details on the 
NowCast algorithm and our implementation of it. We also provide examples to 
highlight specific attributes and potential points of confusion in the algorithm.  

This vignette also briefly covers the `monitor_aqi()` function, which uses the 
`monitor_nowcast()` function to convert raw PM2.5 data into NowCast-based AQI 
values.  

# What is NowCast?

NowCast is an air quality data smoothing algorithm that puts an emphasis on 
recent values when measurements are unstable, and approaches a long-term 
(*e.g.* 12-hour) average when measurements are stable.  

The original algorithm, known as the Conroy method, was developed in 2003 to 
make real-time air quality measurements roughly comparable to established 
regulatory air quality health thresholds (*e.g.* 24‐hour PM2.5 standards). 
However, that method was shown to be slow to respond to rapidly changing air 
quality conditions, which, at best, reduced public confidence in disseminated 
AQI values an, at worst, had the potential to adversely affect the health of 
those in high impact areas (*e.g.* near wildfires).  

In response, EPA developed a new method -- the Reff method -- in 2013 to be more 
responsive to rapidly changing air quality conditions. The **AirMonitor**
package provides support for applying the new NowCast algorithm to hourly PM2.5 
data, though theoretically it could be applied to regular-interval time series 
data of any type, including other criteria pollutants.  

Algorithm specifics are provided below.  

Sources:  

* https://en.wikipedia.org/wiki/NowCast_(air_quality_index)
* https://forum.airnowtech.org/t/the-nowcast-for-pm2-5-and-pm10/172

# The NowCast Algorithm (Reff Method)

## NowCast Equation for PM Data (PM~2.5~ or PM~10~)

The NowCast value for a given hour can be calculated as follows:

$$
NowCast = \frac{\sum_{i=1}^{12}{w^{i-1}c_i}}{\sum_{i=1}^{12}w^{i-1}}
$$

where $i$, $w$ and $c$ are as defined in the sections below.

Source:

* https://en.wikipedia.org/wiki/NowCast_(air_quality_index)

## Data Selection

For PM2.5 data, the NowCast algorithm uses hourly averages from the prior 12 
clock hours.

The hourly averages are denoted below by $c_i$, where $i$ represents the number 
of hours before present. For example, $c_1$, $c_2$, $c_3$, $...$, $c_N$ represent 
the hourly averages for the most recent 1, 2, 3, $...$, $N$ hours.  

Source:

* https://forum.airnowtech.org/t/the-nowcast-for-pm2-5-and-pm10/172

## Weight Factor

Using data points as specified above, define the weight factor $w^*$ as follows:

$$
w^* = 1- \frac{c_{max}-c_{min}}{c_{max}} = \frac{c_{min}}{c_{max}}
$$

where
 
 - $c_{max}$ = the highest value in $c_1$, $...$, $c_N$
 - $c_{min}$ = the lowest value in $c_1$, $...$, $c_N$

NowCast-related literature usually gives the equation for $w^*$ in one of two 
basic forms, both of which are included above for reference. Note that both forms are equivalent.

Sources: 

* https://en.wikipedia.org/wiki/NowCast_(air_quality_index)
* https://forum.airnowtech.org/t/the-nowcast-for-pm2-5-and-pm10/172

## Minimum Weight Factor

Before plugging into the NowCast equation, $w*$ is updated to $w$ for PM2.5 and 
PM10 as follows:

$$
w = 
\begin{cases}
w^*         & \text{if} & w^*>\frac{1}{2} \\
\frac{1}{2} & \text{if} & w^*\leq \frac{1}{2}  \\
\end{cases}
$$

Source

* https://forum.airnowtech.org/t/the-nowcast-for-pm2-5-and-pm10/172

## Truncation

Final NowCast values are truncated based on the type of data being processed:

- **PM2.5**: 0.1 µg/m3
- **PM10**: 1 µg/m3

Source:

* https://forum.airnowtech.org/t/the-nowcast-for-pm2-5-and-pm10/172

# Algorithm Expansion

For PM2.5, the NowCast equation can be expanded as follows:

$$
NowCast = \frac
{w^0c_1 + w^1c_2 + w^2c_3 + w^3c_4 + w^4c_5 + w^5c_6 + w^6c_7 + w^7c_8 + w^8c_9 + w^9c_{10} + w^{10}c_{11} + w^{11}c_{12}}
{w^0 + w^1 + w^2 + w^3 + w^4 + w^5 + w^6 + w^7 + w^8 + w^9 + w^{10} + w^{11}}
$$

Note that:

 - $w^0$, $w^1$, $w^2$, $...$ represent $w$ to the power of 0, 1, 2, $...$ 
 - $c_1$, $c_2$, $c_3$, $...$ represent the hourly average for the most recent 1, 2, 3, $...$ hours.  

When written this way, it is easy to see that in the extreme case where $w = 1$ (i.e. if $c_{min} = c_{max}$) the equation above reduces to:

$$
NowCast = \frac{\sum_{i=1}^{12}{c_i}}{12}
$$

which is just a simple 12-hour arithmetic average. Incidentally, all 12 hourly 
averages and the 12-hour average itself would all be equivalent in this case.  

In the case of highly variable PM2.5 data, $w$ would be set to the minimum value 
of $1/2$, and the most recent data would carry the majority of the weight in the 
equation above.

# Implementation Details

## Missing Data

The NowCast algorithm ignores terms corresponding to hours for which a valid 
observation is not available. For example, suppose PM2.5 is invalid for all but 
the first three and last three hours of a 12-hour period. Then the PM2.5 NowCast 
equation takes the following form:

$$
NowCast = \frac
{w^0c_1 + w^1c_2 + w^2c_3 + \color{gray}{\text{[note: middle values ignored]}} + w^9c_{10} + w^{10}c_{11} + w^{11}c_{12}}
{w^0 + w^1 + w^2 + \color{gray}{\text{[note: middle values ignored]}} + w^9 + w^{10} + w^{11}}
$$

Minimum data availability requirements do apply, however. See the following 
section for details.

## Data Availability Requirements

To get a valid NowCast value, the NowCast algorithm simply requires valid data 
for at least two of the three most recent clock hours. This means that a valid 
NowCast value can be calculated from as little as two valid hours, even if 12 
hours are typically used in the calculation.

However, note that https://en.wikipedia.org/wiki/NowCast_(air_quality_index) says:

> Because the most recent hours of data are weighted so heavily in the NowCast 
> when PM levels are changing, EPA does not report the NowCast when data is 
> missing for c~1~ or c~2~.  

While this seems like a reasonable approach, we could not find a source for 
this statement (in the Wikipedia page sources or elsewhere). We take a 
compromise approach  and still allow valid NowCast values to be calculated when 
c~2~ is invalid but return `NA` when c~1~ is invalid.

## Argument: `includeShortTerm`

As mentioned above, the NowCast algorithm only requires two valid hours to 
calculate valid values. Does this mean that the `monitor_nowcast()` function can 
begin reporting valid values after the second hour in the data (assuming the 
first two hours are both valid)?  

We assert that it would be inappropriate to do so, _usually_.  

In most cases a user will have created a _mts_monitor_ object using one of the 
`~_load()`, functions, which return data for a 
specific period of time. Data before this period is not necessarily invalid, it 
was simply not retrieved. But the function itself has no way of knowing whether 
such earlier data exists, so it has no choice but to consider earlier hours 
"invalid". This means that, if followed by-the-book, the NowCast algorithm could 
return different values for a given hour depending on whether or not the earlier 
data had been retrieved. This is not a desirable behavior, so, by default, the 
`monitor_nowcast()` returns invalid NowCast values until the $N$^th^ hour of data.  

However, we provide a manual override in `includeShortTerm = TRUE` which causes 
the function to return valid values as per the bare-bones data availability 
requirements described above, treating the hours before the beginning of the 
data as invalid. Thus, it can return "valid" NowCast values as early as the 
second (valid) hour in the data.  

This argument may be useful (or even appropriate) for datasets where the 
beginning of the data truly corresponds to the beginning of the monitoring, such 
as when a new monitor has just been installed. In this case, data prior to the 
first hour is truly invalid since it does not exist. The argument may also be 
useful for field personnel looking to ensure that their monitor has been 
successfully plugged in to the data processing pipeline, even if the values 
themselves are not truly representative of the data for the past $N$ hours.  

## Negative Values

In the NowCast literature we find no mention of negative values, which, while 
aphysical, are common in air quality monitoring data. Thus, 
we do not adjust negative values up to zero in the `monitor_nowcast()` function 
itself.

Negative values should always be handled prior to converting  raw data to 
NowCast or AQI values. All of the `~_load()` functions handle negative values 
by _lifting_ them to zero as part of the data loading process.

## Argument: `version`

The `version` argument sets defaults for the number of hours in the lookback $N$ 
(`numHrs`), the minimum weight factor $c_{min}$ (`weightFactorMin`), and the 
number of digits to which the final data is truncated (`digits`).

* `version='pm25'` (default)
    + `numHrs <- 12`
    + `weightFactorMin <- 0.5`
    + `digits <- 1`
* `version='ozone'`
    + `numHrs <- 8`
    + `weightFactorMin <- NA`
    + `digits <- 3`
* `version='pmAsian'`
    + `numHrs <- 3`
    + `weightFactorMin <- 0.1`
    + `digits <- 1`

The default setting is `version = 'pm25'` since this is the parameter most 
commonly stored in _mts_monitor_ objects.  

`version = 'ozone'` supports the O3 NowCast as described above.

`version='pmAsian'` supports an alternative shorter-term NowCast as proposed here:
https://aqicn.org/faq/2015-03-15/air-quality-nowcast-a-beginners-guide/

Although the NowCast algorithm itself supports PM10, we do not currently provide 
functionality for this parameter in the `monitor_nowcast()` function.

# NowCast and AQI Calculations

The EPA uses an [Air Quality Index](https://www.airnow.gov/aqi/aqi-basics/) to put
different pollutants on the same scale. From their site:

> Think of the AQI as a yardstick that runs from 0 to 500. The higher the AQI 
> value, the greater the level of air pollution and the greater the health 
> concern. For example, an AQI value of 50 represents good air quality with 
> little potential to affect public health, while an AQI value over 300 
> represents hazardous air quality.

We provide the `monitor_aqi()` function to convert the PM2.5 data in a 
_mts_monitor_ object into  NowCast-based AQI values in the 0-500 range.

# Examples

The following examples demonstrate the functionality of `monitor_nowcast()` and 
specifics of its implementation.

## Setup

For the following examples we will use the Northwest Megafires data from the 
**AirMonitor** package. In particular, we will look at PM2.5 data from Omak, WA, 
which was heavily impacted by smoke from wildfires during the second half of 
August, 2015:

```{r Omak_nowcast}
library(AirMonitor)

Omak_ID <- "decef259cdefa79f_530470013_04"

# Recipe to obtain monitoring data for Omak, WA
Omak <-
  
  # start with NW Megafires dataset
  NW_Megafires %>%
  
  # select the Omak time series
  monitor_select(Omak_ID) %>%
  
  # filter to include only August, 2015
  monitor_filterDate(20150801, 20150901)
```

```{r Omak_nowcast_plot, echo = FALSE}
monitor_timeseriesPlot(Omak, type = 'l', lwd = 2)
monitor_timeseriesPlot(
  monitor_nowcast(Omak), 
  add = TRUE, 
  type = 'l', 
  lwd = 2, 
  col = 'purple',
  main = "Hourly and Nowcast PM2.5 Values\nOmak, Washington; August, 2015"
)
addAQILines()
addAQILegend('topleft', lwd = 1, pch = NULL)
legend(
  'topright', 
  lwd = 2, 
  col = c('black', 'purple'), 
  legend = c('hourly', 'nowcast')
)
```

## Example 1: Basic Formula Verification

In the code above we used the `monitor_nowcast()` function to calculate PM2.5 
NowCast values for the Omak _mts_monitor_ object. Let's see if we can verify the 
function's output for a single hour.

Below is the Omak PM2.5 data for the first 12 hours of 8/21/15. Let's see if we 
can verify the accuracy of the NowCast value for the last hour in this series, 
8/21/15 Hour 11.

```{r}
Omak_1 <-
  Omak %>%
  monitor_filterDatetime(2015082100, 2015082112, timezone = "UTC")

print(Omak_1$data)
```

First we'll pull out just the values themselves, and reverse them so the most 
recent values come first (_i.e._ so the vector represents $c_1$, $c_2$, $...$, $c_N$).

```{r example_1_values}
example_1_values <-
  Omak_1 %>% 
  monitor_getData() %>%
  dplyr::pull(2) %>%
  rev()

print(example_1_values)
```

Per the NowCast algorithm, we define $w*$ as $\frac{c_{min}}{c_{max}}$:

```{r}
(w_star <- min(example_1_values)/max(example_1_values))
```

We now define $w$ based on $w^*$ and the minimum weight factor; recall that 
$w_{min}=\frac{1}{2}$ for PM2.5:

```{r}
(w <- max(1/2, w_star))
```

Thus, the numerator of the NowCast equation for Hour 11

$$
w^0c_1 + w^1c_2 + w^2c_3 + w^3c_4 + w^4c_5 + w^5c_6 + w^6c_7 + w^7c_8 + w^8c_9 + w^9c_{10} + w^{10}c_{11} + w^{11}c_{12}
$$

becomes the following:

$0.5^0 \times 46.3 + 0.5^1 \times 27.4 + 0.5^2 \times 59.8 + 0.5^3 \times 129.2 + 0.5^4 \times 130.6 + 0.5^5 \times 215.4$ +  
$0.5^6 \times 143.2 + 0.5^7 \times 93.7 + 0.5^8 \times 101.8 + 0.5^9 \times 49.3 + 0.5^{10} \times 80.2 + 0.5^{11} \times 123.3$

which we can calculate in R as follows

```{r}
(numerator <- sum(w^(0:11) * example_1_values))
```

Meanwhile, the denominator

$$
w^0 + w^1 + w^2 + w^3 + w^4 + w^5 + w^6 + w^7 + w^8 + w^9 + w^{10} + w^{11}
$$

becomes the following:

$0.5^0 + 0.5^1 + 0.5^2 + 0.5^3 + 0.5^4 + 0.5^5 + 0.5^6 + 0.5^7 + 0.5^8 + 0.5^9 + 0.5^{10} + 0.5^{11}$

which we can calculate in R as follows

```{r}
(denominator <- sum(w^(0:11)))
```

Dividing the numerator by the denominator gives a value of

```{r echo = FALSE}
(numerator/denominator)
```

which we truncate to one decimal place for a final NowCast value of 
`r trunc(10*numerator/denominator)/10` for 2015-08-21 11:00 local time.

So how does this compare to our `monitor_nowcast()` output for the same hour.

```{r}
Omak %>%
  monitor_nowcast() %>%
  monitor_filterDatetime(2015082111, 2015082112, timezone = "UTC") %>%
  monitor_getData() %>%
  dplyr::pull(Omak_ID)
```

Right on!

## Example 2: Short Missing Data Period

So we have verified the calculation for a period with 12 valid hours. But how
does the function handle missing data periods? Let's take a look at a short
period of missing data to find out. We will invalidate a single hour and see
how the NowCast algorithm responds.

```{r Omak_missing_data}
# Create Omak_2 with one value missing
Omak_2 <- Omak
Omak_2$data[564, 2] <- NA

Omak_2 %>%
  monitor_filterDatetime(2015082412, 2015082500, timezone = "UTC") %>%
  monitor_getData()
```

Here we see that the monitor data is missing for 18:00. We want to verify that
this hour is properly excluded from the NowCast calculation for subsequent hours.
We also want to ensure that NowCast returns valid values for all hours in the
vicinity, since all hours meet the minimum data availability requirements (i.e.
two of three most recent hours valid).

### Formula Verification

Let's see if we can verify the accuracy of the NowCast value for Hour 23.

We again begin by pulling out our vector of values $c_1$, $c_2$, $...$, $c_N$

```{r example_2_values}
example_2_values <-
  Omak_2 %>% 
  monitor_filterDatetime(2015082412, 2015082500, timezone = "UTC") %>%
  monitor_getData() %>%
  dplyr::pull(2) %>%
  rev()

print(example_2_values)
```

and calculating $w*$ and $w$

```{r}
w_star <- min(example_2_values, na.rm = TRUE)/max(example_2_values, na.rm = TRUE)
(w <- max(1/2, w_star))
```

This time our numerator and denominator should both exclude the 6th term, since
$c_6$ is invalid. So we have

```{r}
validIndexes <- which(!is.na(example_2_values))
numerator <- sum(w^(validIndexes - 1) * example_2_values[validIndexes])
denominator <- sum(w^(validIndexes - 1))
numerator/denominator
```

We truncate the value above to one decimal place for a final NowCast value of
`r trunc(10*numerator/denominator)/10` for 2015-08-24 23:00.

So how does this compare to our `monitor_nowcast()` output for the same hour?

```{r}
Omak_2 %>%
  monitor_nowcast() %>% 
  monitor_filterDatetime(2015082423, 2015082500, timezone = "UTC") %>%
  monitor_getData() %>%
  dplyr::pull(Omak_ID)
```

Excellent!

### Missing Data Verification

A quick look at the monitored data alongside the NowCast data shows that all
NowCast hours are valid except for the c~1~ moment when the monitoring data is
also invalid.

```{r}
tbl <- 
  Omak_2 %>%
  monitor_filterDatetime(2015082412, 2015082500, timezone = "UTC") %>%
  monitor_getData()

tbl$nowcast <-
  Omak_2 %>%
  monitor_nowcast() %>%
  monitor_filterDatetime(2015082412, 2015082500, timezone = "UTC") %>%
  monitor_getData() %>%
  dplyr::pull(Omak_ID)

print(tbl)
```

## Example 3: Long Missing Data Period

We now look at a longer period of missing data, from the day prior on 8/23/15.

```{r Omak_long_missing_data}
# Create Omak_3 with 5 missing values
Omak_3 <- Omak
Omak_3$data[538:542, 2] <- NA

Omak_3 %>%
  monitor_filterDatetime(2015082312, 2015082400, timezone = "UTC") %>%
  monitor_getData()
```

Here we see that the monitored data is missing for Hours 16-20. We again want to
verify that these hours are properly excluded from the NowCast calculation for
subsequent hours. This time we also want to ensure that NowCast returns invalid
values for hours in which the minimum data availability requirements are not met
(i.e. two of three most recent hours valid).

### Formula Verification

Let's see if we can verify the accuracy of the NowCast value for Hour 23.

We again begin by pulling out our vector of values $c_1$, $c_2$, $...$, $c_N$

```{r example_3_values}
example_3_values <-
  Omak_3 %>%
  monitor_filterDatetime(2015082312, 2015082400, timezone = "UTC") %>%
  monitor_getData() %>%
  dplyr::pull(2) %>%
  rev()

print(example_3_values)
```

and calculating $w*$ and $w$

```{r}
w_star <- min(example_3_values, na.rm = TRUE)/max(example_3_values, na.rm = TRUE)
(w <- max(1/2, w_star))
```

This time our numerator and denominator should both exclude the 4th-8th terms,
since $c_4$, $c_5$, $c_6$, $c_7$ and $c_8$ are invalid. So we have

```{r}
validIndexes <- which(!is.na(example_3_values))
numerator <- sum(w^(validIndexes - 1) * example_3_values[validIndexes])
denominator <- sum(w^(validIndexes - 1))
numerator/denominator
```

We truncate the value above to one decimal place for a final NowCast value of
`r trunc(10*numerator/denominator)/10` for 2015-08-23 23:00.

So how does this compare to our `monitor_nowcast()` output for the same hour?

```{r}
Omak_3 %>%
  monitor_nowcast() %>%
  monitor_filterDatetime(2015082323, 2015082400, timezone = "UTC") %>%
  monitor_getData() %>%
  dplyr::pull(Omak_ID)
```

Again, right on!

#### Missing Data Verification

Now we'll look at the monitored data alongside the NowCast data to see how the
validity of the NowCast data was affected by the missing monitored hours.

```{r}
tbl <- 
  Omak_3 %>%
  monitor_filterDatetime(2015082312, 2015082400, timezone = "UTC") %>%
  monitor_getData()

tbl$nowcast <-
  Omak_3 %>%
  monitor_nowcast() %>%
  monitor_filterDatetime(2015082312, 2015082400, timezone = "UTC") %>%
  monitor_getData() %>%
  dplyr::pull(Omak_ID)

print(tbl)
```

There's a lot going on here so let's walk through the data one step at a time.

**Hours 12-15**: NowCast data is valid because the monitored data is valid for
all of the three most recent hours.
Status: <span style="color:limegreen">GOOD</span>

**Hour 16**: This is the first hour for which the monitored data is invalid.
NowCast should also return invalid for this c~1~ timestep.
Status: <span style="color:limegreen">GOOD</span>

**Hour 17**: This NowCast value is invalid because the monitored data is only
valid for one of the three most recent hours (Hour 15).
Status: <span style="color:limegreen">GOOD</span>

**Hours 18-20**: NowCast is invalid since the monitored data is valid for none
of the three most recent hours.
Status: <span style="color:limegreen">GOOD</span>

**Hour 21**: This is the first hour for which the monitored data is valid again.
However, NowCast still returns an invalid value this hour since the monitored
data is only valid for one of the three most recent hours (Hour 21).
Status: <span style="color:limegreen">GOOD</span>

**Hour 22**: This is the first hour for which the NowCast data is valid again.
This is because the monitored data is again valid for two of the three most
recent hours (Hours 21 and 22).
Status: <span style="color:limegreen">GOOD</span>

**Hour 23**: NowCast data is valid because the monitored data is valid for all
of the three most recent hours.
Status: <span style="color:limegreen">GOOD</span>

So, it appears the `monitor_nowcast()` returns data that has been correctly
validated according to the NowCast algorithm and associated data availability requirements.

#### Does "most recent" include "current"?

A subtle point about the `monitor_nowcast()` function which falls out of the
example above: **the "current" hour is considered to be a part of the three
"most recent clock hours"**. This may seem strange at first, but to understand
why we chose this approach, one must think about how the data is captured and
how measurement timestamps correspond to the period they actually represent.

For particle pollution, measurements typically represent the mass of particles
that accumulate on a filter during a set period of time, e.g. the past hour.
Measurements come in at the end of the hour, but these measurements aren't
necessarily representative of conditions at the exact times of the measurements.
Instead, they are actually representative of concentrations **during the previous
clock hour**. For example, if an hourly measurement comes in at 12:00, the
measurement is actually representative of data during Hour 11, **NOT** Hour 12.
So we call the 12:00 measurement the Hour 11 data point.

This is really just a labeling convention, and it is no different from how we
treat other datasets. For example, suppose you wanted to take an average of the
low temperatures for every day in July. You would have to wait until August 1st
to do so, since the temperature might still be dropping at the end of the day on
July 31st. But does this mean you would call this value the average low
temperature for August, since it wasn't calculated until August? Of course not:
the data belongs to July, even if it couldn't be calculated until August. It is
no different with particle pollution: the Hour 11 data belongs to Hour 11, even
if it wasn't calculated until 12:00. As a result of this convention, timestamps
are usually an entire hour (or more) earlier than the time the measurements were
actually taken (exact differences depend on several factors).

Another reason for including the "current" hour in the NowCast "three most recent
hours" is for speed of updates. Suppose it is 12:04, and a measurement just came
in at 12:00 (the Hour 11 measurement). It would be inappropriate to wait until
13:00 to calculate the updated NowCast value. For this reason, we calculate
NowCast values using the monitored data for the "current" hour and the $N-1$
prior hours.

### Example 4: `includeShortTerm` argument

For our final NowCast example we will explore the `includeShortTerm` argument,
including a demonstration of why it defaults to `FALSE`.

Suppose we wanted to know the NowCast values for Omak for the first half of
8/25/15, since concentrations were extremely high at the time. Not thinking
about the details of the NowCast algorithm, one might proceed by filtering 
in time before creating the nowcast:

```{r}
Omak %>%
  monitor_filterDatetime(2015082500, 2015082600, timezone = "UTC") %>%
  monitor_nowcast() %>%
  monitor_getData()
```

Unfortunately, the first 11 hours end up as invalid. But you read in the documentation
that `includeShortTerm = TRUE` will return values for the second valid hour
onwards. So you try again, this time setting `includeShortTerm = TRUE`. Let's
check it out.

```{r}
Omak %>%
  monitor_filterDatetime(2015082500, 2015082600, timezone = "UTC") %>%
  monitor_nowcast(includeShortTerm = TRUE) %>%
  monitor_getData()
```

Nice. We were able to get NowCast values for all but the first hour!

But there's a problem lurking. Setting `includeShortTerm = TRUE` caused the
`monitor_nowcast()` function to treat hours prior to Hour 0 as invalid.
**It has no idea that valid data is available for this period.**

As an experiment, let's see how the NowCast values would look if we calculated
them from a larger dataset that includes the data prior to 8/25/15 Hour 0.

```{r}
tbl <-
  Omak %>%
  # NowCast first
  monitor_nowcast %>%
  # then restrict times
  monitor_filterDatetime(2015082500, 2015082600, timezone = "UTC") %>%
  monitor_getData()

tbl$shortTerm_T <-
  Omak %>% 
  # restrict times first
  monitor_filterDatetime(2015082500, 2015082600, timezone = "UTC") %>%
  # then NowCast
  monitor_nowcast(includeShortTerm = TRUE) %>%
  monitor_getData() %>%
  dplyr::pull(Omak_ID)

print(tbl)
```

While not too extreme, we see differences in the NowCast values calculated by
the different approaches, even as late as the 11th hour of the period. As
expected, the values match from the 12th hour on.

The discrepancies displayed above are why, by default, `includeShortTerm = FALSE`.
It should be used only when necessary, and with an understanding that the first
$N$ hours' values might not necessarily be true NowCast values.

If it exists, we recommend always grabbing an extra day of data at the beginning
of the period if you think you might want to calculate NowCast values on a dataset.

----

Hopefully, we have thoroughly addressed any questions about the **AirMonitor**
implementation of the NowCast algorithm.
