---
title: "Example: Save Data as CSV"
author: "Jonathan Callahan"
date: "2022-03-29"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example: Save Data as CSV}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 5)
```

> How can I save data from temporary monitors to a CSV file?

If you have **AirMonitor** installed, you should be able to copy and paste this
code into the RStudio console.

The coding style uses the "pipe" operator, `%>%`, which uses the output of the 
preceding function as the first argument of the next function. Package functions 
are specifically designed to work well in this manner, encouraging **_readable 
and understandable code_**.

Think of each chunk as a _recipe_ that begins with what you want to make and
is followed by the steps needed to make it.

Enjoy!

```{r workedExample, eval = FALSE}
library(AirMonitor)

# AIRSIS in California
#  - start with all AIRSIS monitors in 2019
#  - subset for those where stateCode is one of "CA"
airsis_ca <- 
  airsis_loadAnnual(2019) %>%
  monitor_filter(stateCode == "CA")

# Interactive map to pick a monitor
monitor_leaflet(airsis_ca)

# Select a single monitor by deviceDeploymentID
#  - start with airsis_ca
#  - subset for the "Mariposa" monitorID
Mariposa <-
  airsis_ca %>%
  monitor_select("aa0cee2d9f2085f8_mariposa.1000")

# Interactive graph to pick some time limits
monitor_dygraph(Mariposa)

# Trim empty periods from the beginning and end of this time series
#  - start with Mariposa
#  - trim periods with missing data
Mariposa <- 
  Mariposa %>%
  monitor_trimDate()

monitor_timeRange(Mariposa)

# Subset this time series to October through December
#  - start with Mariposa
#  - subset based on a date range
Mariposa <- 
  Mariposa %>%
  monitor_filterDate(20191008, 20200101)

monitor_timeRange(Mariposa)

# A quick plot for October through December
monitor_timeseriesPlot(
  Mariposa,
  addAQI = TRUE,
  main = "Mariposa 2019"
)
addAQILegend("topright")

# Dump out a meta/data combined CSV file for a subset of Mariposa
Mariposa %>%
  monitor_filterDate(20191101, 20191102) %>%
  monitor_toCSV() %>%
  cat(file = "")
  
# Review the metadata at the command line
dplyr::glimpse(Mariposa$meta)

# Alternatively, View the data (or metadata) in the RStudio table viewer:
View(Mariposa$data)

# Dump out airsis_ca meta to a CSV file
airsis_ca %>%
  monitor_toCSV(includeData = FALSE) %>%
  cat(file = file.path(tempdir(), "airsis_CA_meta.csv"))

# Dump out airsis_ca data to a CSV file
airsis_ca %>%
  monitor_toCSV(includeMeta = FALSE) %>%
  cat(file = file.path(tempdir(), "airsis_CA_data.csv"))

# Alternatively, View() the metadata and data in RStudio:
View(airsis_ca$meta)
View(airsis_ca$data)

# ==============================================================================

# Everything above also applies to monitoring data from other sources.
# Just start with one of the following:

# All of WRCC for 2019
#wrcc_loadAnnual(2019)

# All of AirNow for 2019
#airnow_loadAnnual(2019)
```

Finally, to emphasize what can be done with pipelines, the following calculates
daily mean timeseries for each California monitor and displays the `data` 
dataframe in the RStudio viewer:

```{r recipe, eval = FALSE}
airsis_loadAnnual(2019) %>%
  monitor_filter(stateCode == "CA") %>%
  monitor_filterDate(20191001, 20201101, timezone = "America/Los_Angeles") %>%
  monitor_trimDate(timezone = "America/Los_Angeles") %>%
  monitor_dailyStatistic(mean) %>% 
  monitor_mutate(round, 1) %>%
  monitor_getData() %>%
  View()
```
